Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-07-25 submission/manuscript.Rmd
add methods

# Experimental Procedures

@@

In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity, resulting in substantial misexpression of tissue-enriched genes within the adult _Kdm5c_-KO hippocampus and amygdala. We defined tissue-enriched genes based on the criteria used in a previously published dataset of 17 C57/Bl6J mouse tissues[@liComprehensiveMouseTranscriptomic2017], which defined genes as tissue-enriched if they had more than 4-fold higher expression when compared to any other tissue. In addition to testis genes identified previously[@scandagliaLossKdm5cCauses2017], we found significant enrichment of muscle, liver, and even ovary-biased genes aberrantly expressed within the male _Kdm5c_-KO brain. It is currently unclear what the impact of these tissue-enriched genes are upon _Kdm5c_-KO brain function and if they ultimately contribute to MRXSCJ-related phenotypes. While the majority of tissue-enriched DEGs were testis and ovary genes with no known brain functions, select liver and muscle-biased DEGs do have known roles in the wild-type brain. For example, _Apolipoprotein C-I (Apoc1)_ is a lipid transport gene highly enriched within the liver[XXX], but is lowly expressed in the wild-type brain. Intriguingly, _Apoc1_ overexpression in the mouse brain can impair learning and memory[@abildayevaHumanApolipoproteinCI2008] and is implicated in Alzheimer's disease in humans[@leducAPOECholesterolHomeostasis2010]. Our results suggest KDM5C fine-tunes the expression of tissue-enriched genes like _Apoc1_ to match the level requried for proper brain function. However, futher studies are required to detrmine if these genes contribute to _Kdm5c_-KO neuronal impairments and MRXSCJ pathology.

@@

Dysregulation of _Kdm5c_-KO tissue identity begins during early emrbyogenesis, as germline and other tissue-enriched genes are aberrantly expressed in epiblast-like cells (EpiLCs). _In vivo_, germline genes are typically decommissioned in epiblast stem cells and remain silenced as the epiblast differentiates into the body's somatic tissues[]. However, a small subset of epiblast stem cells will receive signals to reactivate germline gene expression to become the primoridal germ cells (PGCs) that will ultimately form the mature germline[]. This process can be mimicked _in vitro_ by differentiating EpiLCs into primordial germ cell-like cells (PGCLCs)[XXX]. Thus, misexpression of germline genes in _Kdm5c_-KO EpiLCs might suggest they are progressing beyond EpiLC differentiation and becoming PGCLCs. Yet, _Kdm5c_-KO EpiLCs had proper expression of EpiLC marker genes for primed pluripotency and we observed no difference in cellular morphology during ESC to EpiLC differentiation. Furthermore, we saw no significant change in _Otx2_ expression, an epiblast stem cell marker that is known to repress epiblast and EpiLC differentiation into PGC and PGCLCs[@zhangOTX2RestrictsEntry2018]. This, in conjunction with _Kdm5c_-KO mice being viable, suggests germline gene expression is occuring ectopically in conjunction with typical developmental programs, rather than a complete shift to germline identity.

@@

>>>>

@@

In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity, resulting in substantial misexpression of tissue-enriched genes within the adult _Kdm5c_-KO hippocampus and amygdala. We defined tissue-enriched genes based on the criteria used in a previously published dataset of 17 C57/Bl6J mouse tissues[@liComprehensiveMouseTranscriptomic2017], which defined genes as tissue-enriched if they had more than 4-fold higher expression when compared to any other tissue. In addition to testis genes identified previously[@scandagliaLossKdm5cCauses2017], we found significant enrichment of muscle, liver, and even ovary-biased genes aberrantly expressed within the male _Kdm5c_-KO brain. It is currently unclear what the impact of these tissue-enriched genes are upon _Kdm5c_-KO brain function and if they ultimately contribute to MRXSCJ-related phenotypes. While the majority of tissue-enriched DEGs were testis and ovary genes with no known brain functions, select liver and muscle-biased DEGs do have known roles in the wild-type brain. For example, _Apolipoprotein C-I (Apoc1)_ is a lipid transport gene highly enriched within the liver, but is also lowly expressed in the wild-type brain. Intriguingly, _Apoc1_ overexpression in the mouse brain can impair learning and memory[@abildayevaHumanApolipoproteinCI2008] and is implicated in Alzheimer's disease in humans[@leducAPOECholesterolHomeostasis2010]. Our results suggest KDM5C fine-tunes the expression of tissue-enriched genes like _Apoc1_ to match the level requried for proper brain function. However, futher studies are required to detrmine if these genes contribute to _Kdm5c_-KO neuronal impairments and MRXSCJ pathology.

@@

Dysregulation of _Kdm5c_-KO tissue identity begins during early emrbyogenesis, as germline and other tissue-enriched genes are aberrantly expressed in epiblast-like cells (EpiLCs). _In vivo_, germline genes are typically decommissioned in epiblast stem cells and remain silenced as the epiblast differentiates into the body's somatic tissues[@bardotMouseGastrulationCoordination2020]. However, a small subset of epiblast stem cells will receive signals to reactivate germline gene expression to become the primoridal germ cells (PGCs) that will ultimately form the mature germline[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014]. This process can be mimicked _in vitro_ by differentiating EpiLCs into primordial germ cell-like cells (PGCLCs)[@hayashiReconstitutionMouseGerm2011]. Thus, misexpression of germline genes in _Kdm5c_-KO EpiLCs might suggest they are progressing beyond EpiLC differentiation and becoming PGCLCs. Yet, _Kdm5c_-KO EpiLCs had proper expression of EpiLC marker genes for primed pluripotency and we observed no difference in cellular morphology during ESC to EpiLC differentiation. Furthermore, we saw no significant change in _Otx2_ expression, an epiblast stem cell marker that is known to repress epiblast and EpiLC differentiation into PGC and PGCLCs[@zhangOTX2RestrictsEntry2018]. This, in conjunction with _Kdm5c_-KO mice being viable, suggests germline gene expression is occuring ectopically in conjunction with typical developmental programs, rather than a complete shift to germline identity.

@@

# Materials and Methods

## Classifying tissue-enriched and germline-enriched genes
Tissue-enriched differentially expressd genes were determined by their classification in a previously published list of genes enriched in 17 male and female mouse tissues[@liComprehensiveMouseTranscriptomic2017]. This study defined expression as greather than 1 Fragments Per Kilobase of transcript per Million mapped read (FPKM) and tissue enrichment as at least 4-fold higher expression than any other tissue.

We curated a list of germline-enriched genes using an RNA-seq dataset from wild-type and germline-depleted (Kit^W/Wv^) male and female mouse embryos from embryonic day 12, 14, and 16 [@sohGeneRegulatoryProgram2015], as well as adult male testes[@muellerIndependentSpecializationHuman2013]. Germline-enriched genes met the following criteria: 1) their expression is greater than 1 FPKM in wild-type germline 2) their expression in any wild-type somatic tissues[@liComprehensiveMouseTranscriptomic2017] does not exceed 20% of maximum expression in wild-type germline, and 3) their expression in the germ cell-depleted (Kit^W/Wv^) germline, for any sex or time point, does not exceed 20% of maximum expression in wild-type germline.

## Cell culture
We utilized our previously established cultures of male wild-type and _Kdm5c_ mutant embryonic stem cells[@samantaActivationXistEvolutionarily2022]. Sex was confirmed by genotyping _Uba1_ on the X and Y chromosomes with the following primers: 5'-TGGATGGTGTGGCCAATG-3', 5'-CACCTGCACGTTGCCCTT-3'. Deletion of _Kdm5c_ was confirmed through the primers 5'-ATGCCCATATTAAGAGTCCCTG-3', 5'-TCTGCCTTGATGGGACTGTT-3', and 5'-GGTTCTCAACACTCACATAGTG-3'.

Emrbyonic stem cells (ESCs) were initially cultured in primed ESC (pESC) media consisting of KnockOut DMEM (Gibco#10829–018), fetal bovine serum (Gibco#A5209501), KnockOut serum replacement (Invitrogen#10828–028), Glutamax (Gibco#35050-061), Anti-Anti (Gibco#15240-062), MEM Non-essential amino acids (Gibco#11140-050), and beta-mercaptoethanol (Sigma#M7522). They were then transitioned into ground-state "naive" ESCs (nESCs) by culturing for four passages in N2B27 media containing DMEM/F12 (Gibco#11330–032), Neurobasal media (Gibco#21103–049), Gluamax, Anti-Anti, N2 supplement (Invitrogen#17502048), and B27 supplement without vitamin A (Invitrogen#12587-010), and beta-mercaptoethanol. Both pESC and nESC media were supplemented with the GSK3 inhibitor CHIR99021 (Sigma #SML1046-5MG), the MEK inhibitor PD0325901 (Sigma #PZ0162-5MG), and leukemia inhibitory factor (LIF, Millipore#ESG1107).

nESCs were differentiated into epiblast-like cells (EpiLCs, 48 hours) and extendend EpiLCs (exEpiLCs, 96 hours) by culturing in N2B27 media containing DMEM/F12, Neurobasal media, Gluamax, Anti-Anti, N2 supplement, B27 supplement (Invitrogen#17504044), beta-mercaptoethanol, fibroblast growth factor 2 (FGF2, R&D Biotechne 233-FB), and activin A (R&D Biotechne 338AC050CF).

Immunocytochemistry of DAZL in EpiLCs was performed by washing cells thrice with phosphobuffered saline (PBS), fixing cells in 4% paraformaldehyde, washing thrice in PBS, blocking in PBS containing 0.3% Triton X-100, and 5% fetal bovine serum for 1 hour, washing thrice with PBS, and incubating in primary antibody (Rabbit anti DAZL, abcam ab34139, 1:200) in the blocking solution overnight at 4C. The next day cells were rinsed thrice with PBS, incubated in secondary antibody (Alexafluor 488 Invitrogen #710369, 1:1000) in blocking buffer, rinsed thrice in PBS, and then imaged. Images were taken blinded for genotype, chosen based on similar levels of DAPI signal, and then quantified via ImageJ.

## RNA sequencing 
### Published datasets
All published datasets are available at the Gene Expression Omnibus (GEO) https://www.ncbi.nlm.nih.gov/geo. Previously published RNA sequencing datasets analyzed in this study included the male wild-type and _Kdm5c_-KO adult amygdala and hippocampus[@vallianatosMutuallySuppressiveRoles2020] (available at GEO: GSE127722) and male wild-type and _Kdm5c_-KO EpiLCs[@samantaActivationXistEvolutionarily2022] (available at GSE: GSE96797).

### Alignment and analysis
After ensuring read quality via FastQC (v0.11.8), reads were then mapped to the mm10 _Mus musculus_ genome (Gencode) using STAR (v2.5.3a), during which we removed duplicates and kept only uniquely mapped reads. Count files were generated by FeatureCounts (Subread v1.5.0), and BAM files were converted to bigwigs using deeptools (v3.1.3) and visualized by the UCSC genome browser. RStudio (v3.6.0) was then used to analyze counts files by DESeq2 (v1.26.0)[@loveModeratedEstimationFold2014a] to identify differentially expressed genes (DEGs) with a q-value (p-adjusted via FDR/Benjamini–Hochberg correction) less than 0.1 and a log2 fold change greater than 0.5. For all DESeq2 analyses, log2 fold changes were calculated with lfcShrink using the ashr package[@stephensFalseDiscoveryRates2016]. MA-plots were generated by ggpubr (v0.4.0), and Eulerr diagrams were generated by eulerr (v6.1.1). Boxplots and scatterplots were generated by ggpubr (v0.6.0) and ggplot2 (v3.3.2). Heatmaps of gene expression were generated using the base R functions scale and hclust and visualized using the R package ComplexHeatmap (v2.12.1). The Upset plot was generated via the package UpSetR (v1.4.0)[@conwayUpSetRPackageVisualization2017]. Gene ontology (GO) analyses were performed by the R package enrichPlot (v1.16.2) using the biological processes setting.

## Chromatin immunoprecipitation followed by DNA sequecing (ChIP-seq)
We analyzed our previously published KDM5C ChIP-seq experiments in primary neuron cultures from the cortex and hippocampus[@iwaseMouseModelXlinked2016] (available at GEO: GSE61036) and EpiLCs[@samantaActivationXistEvolutionarily2022] (available at GEO: GSE96797). ChIP-seq of histone 3 lysine 4 dimethylation in male EpiLCs[@samantaActivationXistEvolutionarily2022] is also available at GEO: GSE96797. ChIP-seq of histone 3 lysine 4 trimethylation in the male amygdala[@vallianatosMutuallySuppressiveRoles2020] are avilable at GEO: GSE127817.

ChIP-seq reads were aligned to mm10 using Bowtie1 (v1.1.2) allowing up to two mismatches. Only uniquely mapped reads were used for analysis. Peaks were called using MACS2 software (v2.1.0.20140616) using input BAM files for normalization, with filters for a q-value < 0.1 and a fold enrichment > 1. Common peak sets were obtained in R via DiffBind (v3.6.5), and count tables for the common peaks were generated with the Bedtools multicov command. We removed “black-listed” genomic regions that often give aberrant signals. Peak proximity to genome annotations was determined by ChIPSeeker (v1.32.1). Enriched motifs were identified using HOMER[@heinzSimpleCombinationsLineageDetermining2010]. Gene ontology (GO) analyses were performed by the R package enrichPlot (v1.16.2) using the biological processes setting. Average binding across the genome was visualized using deeptools (v3.1.3). Bigwigs were visualized using the UCSC genome browser.

## Whole genome bisulfite sequencing (WGBS)

## Data availability

# Acknowledgements
